Analytic choices for analyzing multidimensional behavior: Many analysts test hypotheses about human speech

### 2021-03-29 RR_manuscript/Draft_RR.Rmd
:memo: - edits to RR (in progress)

---
title: "Analytic choices for analyzing multidimensional behavior: Many analysts test hypotheses about human speech"
runninghead: "Many analysts, one speech data set"
author:
- name: Stefano Coretta*
  num: 1
- name: Timo Roettger
  num: 2
- name: Joseph V. Casillas
  num: 3
address:
- num: 1
  org: Institute of Phonetics and Speech Processing, Ludwig-Maximilian University Munich, Germany.
- num: 2
  org: Department of Applied Things, University B, City B, Country B
- num: 3
  org: Very Important Stuff Committee, Institute C, City C, Country C
corrauth: "Stefano coretta."
email: "s.coretta@lmu.de"
abstract: "TBA"
keywords: "crowdsourcing science, data analysis, scientific transparency, speech, acoustic analysis"
classoption:
  - Review
  - times
bibliography: references
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
    includes:
      in_header: preamble.tex
---

@@

In this article, we will investigate the variability in analytic choices when many analyst teams analyze the same speech production data set, a process that involves both decisions regarding the operationalization of a complex  signal and decisions regarding the statistical analysis.
<!-- TODOLATER: include number of teams when known -->
Specifically, we will report the impact of analytic decision making on research results obtained by a minimum of 12 teams who will gain access to the same set of acoustic recordings in order to answer the same research question.
<!-- TODO: check tense throughout, should be future tense -->

@@

Data analysis comes with many decisions, like how to measure a given phenomenon or behavior, which data to submit to statistical modeling and which to exclude in the final analysis, or what inferential decision making procedure to apply.
This can be problematic because humans show cognitive biases that can lead to erroneous inferences.
Humans filter the world in irrational ways [e.g., @tversky1974judgment], seeing coherent patterns in randomness [@brugger2001], convincing themselves of the validity of prior expectations ["I knew it", @nickerson1998confirmation], and perceiving events as being plausible in hindsight ["I knew it all along", @fischhoff1975hindsight].
In conjunction with an academic incentive system that rewards certain discovery processes more than others [@sterling1959publication; @koole2012rewarding], we often find ourselves exploring many possible analytic pipelines, but only reporting a select few.

@@

Several other recent studies corroborated this analytic flexibility across different disciplines.
Both @dutilh2019 ad @starns2019assessing investigated analysts' choices when inferring theoretical constructs based on the same data set using computational models.

Both studies revealed vastly different modeling strategies, even though scientific conclusions were similar across analysis teams [see also @Parker2020 on analytic flexibility in ecology and @botvinik-nezer2020 on neuroimaging data].
@bastiaansen2020 crowdsourced clinical recommendations based on analyses of an individual patient.
Results suggests that analysts differed substantially regarding decisions related to both the statistical analysis of the data and the theoretical rationale behind interpreting the statistical results.

While these studies attested a large degree of analytic flexibility with possibly impactful consequences, they only investigated analytic decisions related to inferential statistics or the architecture of computational models.
In these studies the data sets were fixed and neither data collection nor measurement could be changed.
However, in many fields the primary raw data are complex signals that need to be operationalized relative to a theoretically motivated research question.
This is especially true in the social sciences, where the phenomenon under investigation corresponds to both observable and unobservable human behavior.
Decisions about how to measure a theoretical construct related to such behavior or its underlying cognitive processes might interact with downstream decisions about statistical modeling and vice versa.

For instance, @flake2020 discuss the cascading impact that different practices can have on psychometric research.
The authors highlight, among others, the following degrees of freedom in the choice and development of measures: definition of the theoretical construct, justification of the selected measure, description of the measure and of how it maps onto the construct, response coding and related transformations, and post-hoc modifications to the chosen measure.
Taken together, these aspects alone exponentially increase the combinations of possible analytic choices, and hence flexibility in research outcomes.

In social sciences concerned with communication, human behavior is sometimes measured as a complex visual or acoustic signal.
This added complexity introduces further fork junctions in the research pipe-line, and in turn further increases analytic flexibility.
The present study looks at experimentally elicited speech production data in order to understand how this flexibility manifests itself in a scenario where complex decision procedures are involved in the operationalization and measuring of complex signals.

@@

Speech is a complex signal that is characterized by structurally different acoustic landmarks distributed throughout different temporal domains.
Thus, choosing how to measure a phenomenon of interest is an important and non-trivial analytic decision.
Take for example the following sentence in (1).

@@

Depending on the speaker's intention, this sentence can be said in different ways.
If, for instance, the speaker is exhausted by all their meetings, the speaker might acoustically highlight the word *another* or *meeting* to contrast it with other, more pleasant activities.
If, on the other hand, the speaker is just tired of video conferences, as opposed to say face-to-face meetings, they might acoustically highlight the word *Zoom*.

If we decide to compare the speech signal associated with these two intentions, how can we quantify the difference between them?
Given their physical manifestation (speech), what do we measure and how do we measure it?
Because of the continuous and transient nature of speech, identifying speech parameters and temporal domains within which to measure those parameters becomes a non-trivial task.
Utterances stretch over several thousand milliseconds and contain different levels of linguistically relevant units such as phrases, words, syllables, and individual sounds.
The researcher is thus confronted with a considerable number of parameters and combinations thereof to choose from.

@@

The multiplicity of phonetic cues grows exponentially if we look at larger temporal domains as is the case for suprasegmental aspects of speech.
For example, studies investigating acoustic correlates of word stress (e.g. the difference between *ínsight* and *incíte*) have been using a wide variety of measurements, including temporal characteristics (duration of certain segments or sub-segmental intervals), spectral characteristics (intensity, formants, and spectral tilt), and measurements related to fundamental frequency (f0) [e.g., @gordon2017acoustic].
Moving on to the expression of higher-level functions, like information structure and discourse pragmatics, relevant acoustic cues can be distributed throughout even larger domains, such as phrases and whole utterances [e.g., @ladd2008intonational].
Differences in position, shape, and alignment of pitch modulations over multiple locations within a sentence are correlated with differences in discourse functions (e.g. Niebuhr et al., 2011).
The latter can also be expressed by global vs local pitch modulations [@heuven2002], as well as acoustic information within the temporal or spectral domain [e.g., @van2005speech].
Extra-linguistic information, like the speaker's intentions, levels of arousal or social identity, are also conveyed by broad-domain parameters, such as voice quality, rhythm, and pitch [@foulkes2006; @ogden2004; @white2009].

In short, when testing hypotheses on speech production data, researchers are faced with many choices and possibilities.
The larger the functional domain (e.g. segments vs words vs utterances), the higher the number of conceivable operationalizations.
For example, several decisions have to be made when comparing the two realization of the sentence in (1), one of which is intended to signal emphasis on *another* and one of which emphasizes *zoom*.

@@

(2a) I can't bear *ANOTHER* meeting on zoom.

@@

Do we compare only the word *another* in (2a) and (2b), or also the word *zoom*?
Do we measure utterance-wide acoustic profiles, whole words, or just the stressed syllables?
Do we average across the chosen time domain or do we measure a specific point in time?
Do we measure fundamental frequency, intensity, or something else?

When looking at phrase-level temporal domains, the number of possible analytic pipelines increases exponentially.
This plurality of analytic paths is illustrated in Figure \@ref(fig:forkingPaths).
Even if we have decided to compare fundamental frequency (f0) of only the word *another* across the two utterances, there are still many choices to be made, all of which need to be justified.
For example, we could measure f0 at specific points in time like the onset of the window, the offset, or the midpoint.
We could also measure the value or time of the minimum or maximum f0 value.
We could summarize f0 across the entire window and extract the mean, median or standard deviation of f0, all of which have been used to analyze speech data [see @gordon2017acoustic].
And the garden of forking paths does not stop here.
Figure \@ref(fig:forkingPaths), illustrates one specific procedure to automatically calculate f0.
However, another options would be to filter the audio signal, to smooth the extracted f0 signal, to remove values that substantially deviate from surrounding values or expectations, either manually or automatically, and so on.

<!-- FIGURE -->
(ref:forkingPaths) Illustrations of the analytic flexibility associated with acoustic analyses. (A) an example of multiple possible and justifiable decisions when comparing two utterances; (B) waveform and fundamental frequency (f0) track of two instances of utterances 2a and 2b. The words "another is highlighted by the green box"; (C) spectrogram and f0 track of the word another, exemplifying different operationalizations of differences in pitch.

@@

These decisions are intended to be made prior to any statistical analysis, but are at times revised a posteriori (i.e. after data collection and/or preliminary analyses) in light of unforeseen or surprising outcomes.
These myriads of possible decisions are exponentiated by those researcher degrees of freedom related to statistical analysis [e.g. @wicherts2016].
In sum, speech data are complex physical signals that constitutes an as of yet unappreciated amount of analytic flexibility when choosing measures and operationalizing underlying theoretical constructs.
The present paper probes this garden of forking paths in the analysis of speech.
To assess the variability in data analysis pipelines across independent researchers, we will provide analytic teams with an experimentally elicited speech production data set.
The study aimed at investigating whether speakers acoustically modify utterances to signal unexpected referring expressions.
In the following, we will shortly introduce the research question of the original study.

@@

Despite the coherent idea that speakers are both rational and efficient, there is much evidence that speakers are often over-informative: Speakers use referring expressions that are more specific than strictly necessary for the unambiguous identification of the intended referent [@sedivy2003pragmatic; @westerbeek2015stored; @rubio2016redundant], which has been argued to facilitate object identification and making communication between speakers and listeners more efficient [@arts2011overspecification; @paraboni2007generating; @rubio2016redundant].
Recent findings suggest that the utility of a referring expression depends on how good it is for a listener (compared to other referring expressions) to identify a target object.
For example, @degen2020redundancy showed that modifiers that are less typical for a given referent (e.g. a blue banana) are more likely to be used in an over-informative scenario (e.g. when there is just one banana).
This account, however, has mainly focused on content selection [@gatt2013we], i.e. whether a certain referential expression is chosen or not, ignoring the fact that speech communication is much richer.

Even looking at morphosyntactically identical expressions, speakers can modulate utterances via suprasegmental acoustic properties like temporal and spectral modifications of the segments involved [e.g., @ladd2008intonational].
Most prominently, languages use intonation to signal discourse relationships between referents (among other functions).
Intonation marks discourse-relevant referents for being new or given information to guide the listeners’ interpretation of incoming messages.
Beyond structuring information relative to the discourse, a few studies suggested that speakers might use intonation to signal atypical lexical combinations [e.g. dimitrova2008prosodic; dimitrova2009did].
Referential expressions such as *blue banana* were produced with greater prosodic prominence than more typical referent such as *yellow banana*.
If true, these findings are in line with the idea of resource-efficient, rational language users who modulate their speech in order to facilitate listeners' comprehension process.
However, the original conclusions of above studies were based on small (N = 10), unjustified sample sizes.
Moreover, the data is not available, and the statistical analysis used anti-conservative co-variance structures, leaving reason to doubt the original conclusions.

To further illuminate the question of whether speakers modify speech to signal atypical referents, thirty native German speakers were recorded while interacting with a confederate (one of the experimenters) in a referential game.
The participants had to verbally instruct the confederate to select a specified target object out of four object presented on a screen.
Figure \@ref(fig:procedural) shows the experimental procedure time-line.
The subject and confederate were seated at the opposite sides of a table, each facing one of two mirrored computer screens.
The participant and the experimenter could not see each other nor each others' screens.
After a familiarization phase (see XXX), the subject first saw four colored objects in the top left, top right, bottom left, and bottom right corners of the screen.
One of the objects served as the target, another as the competitor, and the remaining two objects served as distractors.
Objects were referred to using noun phrases consisting of an adjective modifier denoting color and a modified object (e.g. *Gelbe Zitrone* 'yellow lemon', *Rote Gurke* 'red cucumber', *Rote Socken* 'red socks').

@@

In the center of the screen, a black cube was displayed, which could be moved by the experimenter.
The participant would read a sentence prompt out loud (*Du sollst den Würfel auf der <COLOR OBJECT> ablegen* 'You have to put the cube on top of the <COLOR OBJECT>') to instruct the experimenter to drag the cube on top of one of the four depicted objects (the *competitor*) using the mouse.
After the experimenter had moved the cube as instructed, the subject would read another sentence prompt (*Und jetzt sollst du den Würfel auf der <COLOR OBJECT> ablegen* 'And now, you have to put the cube on top of the <COLOR OBJECT>') instructing the experimenter to move the cube on top of a different object (the *target*).
The second utterance in the trial was the critical trial for analysis.

The two sentence prompts were used to create a focus contrast between the competitor and the target object.
If the competitor and target objects differed but their color did not (e.g. *yellow banana* vs *yellow tomato*), the noun was discourse-pragmatically focused (the Noun Focus condition, NF).
If the objects were the same but differed in color (e.g. *yellow banana* vs *blue banana*), the color adjective was discourse-pragmatically focused (the Adjective Focus condition, AF).
If both the color and the object differed (e.g. *yellow banana* vs *blue tomato*), the whole noun phrase was discourse-pragmatically focused (the Adjective/Noun Focus condition, ANF).
The NF condition constituted the experimentally relevant condition, while the AF and ANF conditions acted as fillers.
The color-object combinations in the Noun Focus (NF) condition were manipulated with respect to their typicality.
The combinations were either typical (e.g. *orange mandarin*), medium typical (e.g. *greem tomato* ), or atypical (e.g. *yellow cherry*), as established by a norming study that was conducted prior to the production experiment (see HERE^[Temporary link: <https://osf.io/5agn9/?view_only=3429acfd8eae4f0aafe2e17ad5b0983f>]).

Each participant responded to 15 critical trials (NF condition).
Each trial was repeated twice, yielding a total of 30 trials per participant and a grand-total 900 ($15 \times 2 \times 30$ participants) spoken utterances.
All analysis teams will be given access to the entire data set and will be instructed to answer the following research question: *Do speakers acoustically modify utterances to signal atypical word combinations in referring expressions?*

>>>>

---
title: "Analytic choices for analyzing multidimensional behavior: Many analysts test hypotheses about human speech"
runninghead: "Many analysts, one speech data set"
author:
- name: Stefano Coretta*
  num: 1
- name: Timo Roettger
  num: 2
- name: Joseph V. Casillas
  num: 3
address:
- num: 1
  org: Institute of Phonetics and Speech Processing, Ludwig-Maximilian University Munich, Germany.
- num: 2
  org: Department of Applied Things, University B, City B, Country B
- num: 3
  org: Very Important Stuff Committee, Institute C, City C, Country C
corrauth: "Stefano Coretta."
email: "s.coretta@lmu.de"
abstract: "TBA"
keywords: "crowdsourcing science, data analysis, scientific transparency, speech, acoustic analysis"
classoption:
  - Review
  - times
bibliography: references
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
    includes:
      in_header: preamble.tex
---

@@

In this article, we investigate the variability in analytic choices when many analyst teams analyze the same speech production data set, a process that involves both decisions regarding the operationalization of a complex  signal and decisions regarding the statistical analysis.
<!-- TODOLATER: include number of teams when known -->
Specifically, we discuss the impact of analytic decision making on research results obtained by N teams who had access to the same set of acoustic recordings in order to answer the same research question.
<!-- TODO: check tense throughout, should be future tense -->

@@

Data analysis comes with many decisions, like how to measure a given phenomenon or behavior, which data to submit to statistical modeling and which to exclude in the final analysis, or what inferential decision-making procedure to apply.
This can be problematic because humans show cognitive biases that can lead to erroneous inferences.
Humans filter the world in irrational ways [e.g., @tversky1974judgment], seeing coherent patterns in randomness [@brugger2001], convincing themselves of the validity of prior expectations ["I knew it", @nickerson1998confirmation], and perceiving events as being plausible in hindsight ["I knew it all along", @fischhoff1975hindsight].
In conjunction with an academic incentive system that rewards certain discovery processes more than others [@sterling1959publication; @koole2012rewarding], we often find ourselves exploring many possible analytic pipelines, but only reporting a selected few.

@@

Several other recent studies corroborated this analytic flexibility across different disciplines.
@dutilh2019 ad @starns2019assessing investigated analysts' choices when inferring theoretical constructs based on the same data set using computational models.
Both studies revealed vastly different modeling strategies, even though scientific conclusions were similar across analysis teams [see also @Parker2020 on analytic flexibility in ecology; and @botvinik-nezer2020 on neuroimaging data].
@bastiaansen2020 crowdsourced clinical recommendations based on analyses of an individual patient.
Results suggests that analysts differed substantially regarding decisions related to both the statistical analysis of the data and the theoretical rationale behind interpreting the statistical results.

While these studies attested a large degree of analytic flexibility with possibly impactful consequences, they only investigated analytic decisions related to inferential statistics or the architecture of computational models.
In these studies the data sets were fixed and neither data collection nor measurement could be changed.
However, in many fields the primary raw data are complex signals that need to be operationalized relative to a theoretically motivated research question.
This is especially true in the social sciences, where the phenomenon under investigation corresponds to both observable and unobservable human behavior.

Decisions about how to measure a theoretical construct related to human behavior or its underlying cognitive processes might interact with downstream decisions about statistical modeling and vice versa.
For instance, @flake2020 discuss the cascading impact that different practices can have on psychometric research.
The authors highlight, among others, the following degrees of freedom in the choice and development of measures: definition of the theoretical construct, justification of the selected measure, description of the measure and of how it maps onto the construct, response coding and related transformations, and post-hoc modifications to the chosen measure.
Taken together, these aspects alone exponentially increase the combinations of possible analytic choices, and hence flexibility in research outcomes.

In those disciplines concerned with communication, human behavior sometimes corresponds to multidimensional visual or acoustic signals.
The complex nature of this data multiplies the number of possible analytic approaches, thus further increasing analytic flexibility.
The present study looks at experimentally elicited speech production data in order to understand how this flexibility manifests itself in a scenario where several procedures are involved in the operationalization and measuring of complex signals.

@@

Speech is a complex signal that is characterized by structurally different acoustic landmarks distributed throughout different temporal domains.
Thus, choosing how to measure a phenomenon of interest is an important analytic decision.
Take for example the following sentence in (1).

@@

Depending on the speaker's intention, this sentence can be said in different ways.
For instance, if the speaker is exhausted by all their meetings, they might acoustically highlight the word *another* or *meeting* to contrast it with other, more pleasant activities.
If, on the other hand, the speaker is just tired of video conferences, as opposed to say face-to-face meetings, they might acoustically highlight the word *Zoom*.

If we decide to compare the speech signal associated with these two intentions, how can we quantify the difference between them?
In other words, given their physical manifestation (speech), what do we measure and how do we measure it?
Because of the continuous and transient nature of speech, identifying speech parameters and temporal domains within which to measure those parameters becomes a non-trivial task.
Utterances stretch over several thousand milliseconds and contain different levels of linguistically relevant units such as phrases, words, syllables, and individual sounds.
The researcher is thus confronted with a considerable number of parameters and combinations thereof to choose from.

@@

The multiplicity of phonetic cues grows exponentially if we look at larger temporal domains as is the case for suprasegmental aspects of speech.
For example, studies investigating acoustic correlates of word stress (e.g. the difference between *ínsight* and *incíte*) have been using a wide variety of measurements, including temporal characteristics (duration of certain segments or sub-segmental intervals), spectral characteristics (intensity, formants, and spectral tilt), and measurements related to fundamental frequency (f0) [e.g., @gordon2017acoustic].
Moving on to the expression of higher-level functions, like information structure and discourse pragmatics, relevant acoustic cues can be distributed throughout even larger domains, such as phrases and whole utterances [e.g., @ladd2008intonational].
Differences in position, shape, and alignment of pitch modulations over multiple locations within a sentence are correlated with differences in discourse functions (e.g. Niebuhr et al., 2011).
The latter can also be expressed by global vs local pitch modulations [@heuven2002], as well as acoustic information within the temporal or spectral domain [e.g., @van2005speech].
Extra-linguistic information, like the speaker's intentions, levels of emotional arousal or social identity, are also conveyed by broad-domain parameters, such as voice quality, rhythm, and pitch [@foulkes2006; @ogden2004; @white2009].

In short, when testing hypotheses on speech production data, researchers are faced with many choices and possibilities.
The larger the functional domain (e.g. segments vs words vs utterances), the higher the number of conceivable operationalizations.
For example, several decisions have to be made when comparing the two realizations of the sentence in (1), one of which is intended to signal emphasis on *another* and one of which emphasizes *Zoom*.

@@

(2a) I can't bear *ANOTHER* meeting on Zoom.

@@

Do we compare only the word *another* in (2a) and (2b), or also the word *Zoom*?
Do we measure utterance-wide acoustic profiles, whole words, or just the stressed syllables?
Do we average across the chosen time domain or do we measure a specific point in time?
Do we measure fundamental frequency, intensity, or something else?

<!-- FIGURE -->
(ref:forkingPaths) Illustration of the analytic flexibility associated with acoustic analyses. (A) An example of multiple possible and justifiable decisions when comparing two utterances; (B) Waveform and fundamental frequency (f0) track of the utterances *I can't bear ANOTHER meeting on Zoom* and *I can't bear another meeting on ZOOM*. The green boxes mark the word *another* in both sentences; (C) Spectrogram and f0 track of the word *another*, exemplifying possible operationalizations of differences in f0.

@@

When looking at phrase-level temporal domains, the number of alternative analytic pipelines substantially increases.
Figure \@ref(fig:forkingPaths)A shows a typical example of a decision tree which speech researcher are often confronted with.
Each of the four analytic decisions in the example have different possible options.
Here only one particular path has been taken, and a different one would likely produce quite different results.
Once we have decided to compare fundamental frequency of the word *another* across the two utterances, there are still many choices to be made, all of which need to be justified.
As Figures \@ref(fig:forkingPaths)B-C illustrate, we could measure f0 at specific points in time like the onset of the temporal window, the offset, or the midpoint.
We could also measure the value or time of the minimum or maximum f0 value.
We could summarize f0 across the entire window and extract the mean, median or standard deviation of f0, all of which have been used to analyze speech data in previous work [see @gordon2017acoustic].
But the journey in the garden of analytic paths goes on.
Other important operationalization steps could involve filtering the audio signal, smoothing the extracted f0 track, removing values that substantially deviate from surrounding values or expectations, either manually or automatically, and so on.

These decisions are intended to be made prior to any statistical analysis, but are at times revised a posteriori (i.e. after data collection and/or preliminary analyses) in light of unforeseen or surprising outcomes.
These myriads of possible decisions are multiplied by those researcher degrees of freedom related to statistical analysis [e.g. @wicherts2016].
In sum, speech data is made of complex physical signals that generates an as-of-yet unappreciated amount of analytic flexibility in the choice of measures and operationalizations.
The present paper probes this garden of forking paths in the analysis of speech.
To assess the variability in data analysis pipelines across independent researchers, we provided analytic teams with an experimentally elicited speech production data set.
The unpublished study we took the data from aimed at investigating whether speakers acoustically modify utterances to signal unexpected referring expressions.
In the following section we introduce the research question and the experimental procedure of said study, and we describe the dataset as it was given to the analysts.

@@

Despite the coherent idea that speakers are both rational and efficient, there is much evidence that speakers are often over-informative.
Speakers use referring expressions that are more specific than strictly necessary for the unambiguous identification of the intended referent [@sedivy2003pragmatic; @westerbeek2015stored; @rubio2016redundant], which has been argued to facilitate object identification and making communication between speakers and listeners more efficient [@arts2011overspecification; @paraboni2007generating; @rubio2016redundant].
Recent findings suggest that the utility of a referring expression depends on how good it is for a listener (compared to other referring expressions) to identify a target object.
For example, @degen2020redundancy showed that modifiers that are less typical for a given referent (e.g. a blue banana) are more likely to be used in an over-informative scenario (e.g. when there is just one banana).
This account, however, has mainly focused on content selection [@gatt2013we], i.e. whether a certain referential expression is chosen or not, while not taking into consideration the indexical potential of speech.

Even when morphosyntactically identical expressions are involved, speakers can modulate utterances via suprasegmental acoustic properties like temporal and spectral modifications [e.g., @ladd2008intonational].
Most prominently, languages use intonation to signal discourse relationships between referents, among other functions.
Intonation marks discourse-relevant referents for being new or given information, to guide the listeners’ interpretation of incoming messages.
Beyond structuring information relative to the discourse, a few studies suggested that speakers might use intonation to signal atypical lexical combinations [e.g. @dimitrova2008prosodic; @dimitrova2009did].
Referential expressions such as *blue banana* were produced with greater prosodic prominence than more typical referents such as *yellow banana*.
These results are in line with the idea of resource-efficient, rational language users who modulate their speech in order to facilitate listeners' comprehension.
However, the above studies are based on a rather small sample size (10 participants) and on potentially anti-conservative statistical analyses, leaving reason to doubt the generalisability of the studies' conclusions.

To further illuminate the question of whether speakers modify speech to signal atypical referents, and overcome some of the limitations of previous work, thirty native German speakers were recorded in a production study while interacting with a confederate (one of the experimenters) in a referential game.
The participants had to verbally instruct the confederate to select a specified target object out of four objects presented on a screen.
The subject and confederate were seated at the opposite sides of a table, each facing one of two mirrored computer screens.
The participant and the experimenter could not see each other nor each others' screens.
Figure \@ref(fig:procedural) shows the experimental procedure time-line.
After a familiarization phase, the subject first saw four colored objects in the top left, top right, bottom left, and bottom right corners of the screen.
One of the objects served as the target, another as the competitor, and the remaining two objects served as distractors.
Objects were referred to using noun phrases consisting of an adjective modifier denoting color and a modified object (e.g. *Gelbe Zitrone* 'yellow lemon', *Rote Gurke* 'red cucumber', *Rote Socken* 'red socks').

@@

In the center of the screen, a black cube was displayed, which could be moved by the experimenter.
The participant would read a sentence prompt out loud (*Du sollst den Würfel auf der COLOR OBJECT ablegen* 'You have to put the cube on top of the COLOR OBJECT') to instruct the experimenter to drag the cube on top of one of the four depicted objects (the *competitor*) using the mouse.
After the experimenter had moved the cube as instructed, the subject would read another sentence prompt (*Und jetzt sollst du den Würfel auf der COLOR OBJECT ablegen* 'And now, you have to put the cube on top of the COLOR OBJECT') instructing the experimenter to move the cube on top of a different object (the *target*).
The second utterance in the trial was the critical trial for analysis.

The two sentence prompts were used to create a focus contrast between the competitor and the target object.
If the competitor and target objects differ but their color did not (e.g. *yellow banana* vs *yellow tomato*), the narrow focus is on the noun (the Noun Focus condition, NF).
If the objects are the same but differ in color (e.g. *yellow banana* vs *blue banana*), the narrow focus is on the color adjective (the Adjective Focus condition, AF).
If both the color and the object differ (e.g. *yellow banana* vs *blue tomato*), the narrow focus is on the whole noun phrase (the Adjective/Noun Focus condition, ANF).
The NF condition constituted the experimentally relevant condition, while the AF and ANF conditions acted as fillers.
Crucially, the color-object combinations in the Noun Focus (NF) condition were manipulated with respect to their typicality.
The combinations were either typical (e.g. *orange mandarin*), medium typical (e.g. *green tomato*), or atypical (e.g. *yellow cherry*), as established by a norming study that was conducted prior to the production experiment just described.^[A detailed description of the norming study, with a more in-depth description of this production study can be found in `methods_norm_prod.pdf` at (temporary link) <https://osf.io/5agn9/?view_only=3429acfd8eae4f0aafe2e17ad5b0983f>]
Each subject produced 15 critical trials (NF condition).
Each trial was repeated twice, yielding a total of 30 trials per participant and a grand-total of 900 ($15 \times 2 \times 30$ participants) spoken utterances.

For the present study, N analysis teams were given access to the entire data set generated by the production study and were instructed to answer the following research question: *Do speakers acoustically modify utterances to signal atypical word combinations in referring expressions?*
...
